---
title: "for-andy"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)
require('here')
require('kableExtra')
library(kableExtra)
library(tidyverse)
library(jagsUI)
jdat<-readRDS(file=paste0(here::here("jdat-jar.rds")))

meta <- read.csv("temp-meta-table.csv")
names(meta)[1] <-"dat"
```
## data object 
The data object **jdat** is a list, currently, of the BBS data alone. 

Important:: the BBS data in this list DO NOT include grid cells where no BBS data exists. I.e., `r jdat$G.bbs` is the # grid cells with BBS data, but total number of grid cells in our study area is actually `r jdat$G.grid`.

##  Naming conventions:1. "C.xxx":BBS counts
1. "p.xxx":detection / observation process covariates
1. "n.xxx":scalars indicating number of xxx
1. "id.xxx":unique identifiers for xxx
1. others:others named to strive for recognizability (e.g., doy = day of year, julian = julian date with some origin)

## jdat contents
Here are the contents of jdat (for now, for bbs...)
```{r jdat-metadata, echo=TRUE}
# kableExtra::kable(names(jdat))
temp=meta %>% arrange(dat, name)
kable(temp) %>% 
  kableExtra::kable_minimal()
```
