---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dubcorms

<!-- badges: start -->
<!-- badges: end -->

In development package for integrating eBird and BBS data. 

## Installation

You can install the development version of dubcorms from [GitHub](https://github.com/) with:

```{r install}
# install.packages("devtools")
devtools::install_github("trashbirdecology/dubcorms")
```

```{r imp-data-metadata}
require('here')
require('kableExtra')
library(kableExtra)
library(tidyverse)
library(jagsUI)
jdat<-readRDS(file=paste0(here::here("inst/jdat-jar.rds")))

meta <- read.csv("inst/data-meta-table.csv")
names(meta)[1] <-"dat"
```


# About Data Object `jdat`
The data object **jdat** is currently a list comprising the BBS and eBird data for use in JAGS (SUBJECT TO CHANGE) 

Important:: the BBS data in this list DO NOT include grid cells where no BBS data exists. I.e., `r jdat$G.bbs` is the # grid cells with BBS data, but total number of grid cells in our study area is actually `r jdat$G.grid`.

##  Naming conventions:1. "C.xxx":BBS counts
1. "p.xxx":detection / observation process covariates
1. "n.xxx":scalars indicating number of xxx
1. "id.xxx":unique identifiers for xxx
1. others:others named to strive for recognizability (e.g., doy = day of year, julian = julian date with some origin)

## jdat contents
Here are the contents of jdat (for now, for bbs...)
```{r jdat-metadata, echo=TRUE}
# kableExtra::kable(names(jdat))
temp=meta %>% arrange(dat, name)
kable(temp) %>% 
  kableExtra::kable_minimal()
```



