---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/trashbirdecology/dubcorms/workflows/R-CMD-check/badge.svg)](https://github.com/trashbirdecology/dubcorms/actions)
<!-- badges: end -->

# dubcorms
The purpose of this R package (*likely to undergo a name change...*) is three-fold:

  1. provide a (currently) faster alternative to the R package `auk ` for importing and munging the large eBird datasets (*see note)
  1. integrate the BBS and eBird observation datasets for use in JAGS (`rjags`) and `mcgv::jagam()`
  1. run fully Bayesian integrated population models using said data
  
*Note: GH users [\@cboettig](https://github.com/cboettig/) and [\@amstrimas](https://github.com/amstrimas/) are currently developing an `auk` alternative, [`birddb`](https://github.com/cboettig/birddb/). It is likely that, once stable, this R package will use `birddb` for eBird import/manipulation. 



## Installation
Download development version from GitHub with:
```{r results=FALSE, echo=TRUE, warning=FALSE, message=FALSE}
# install.packages("devtools")
devtools::install_github("trashbirdecology/dubcorms")
```

## eBird Data Requirements
Prior to using this package, you must have downloaded eBird data. To request and download eBird observations, visit [the eBird website](https://ebird.org/data/download). Credentials are required, and may take up to a few business days for approval, depending on use case. For more information on the eBird data see the eBird website, or visit [the repository for the Cornell Lab of Ornithology's offical R package for munging eBird data,  `auk`](https://github.com/CornellLabofOrnithology/auk/).

When your account is approved, you will gain access to the [eBird Basic Database (EBD)](https://ebird.org/data/download/ebd). This package requires two components of the EBD to be saved to local file:

1. the **observations** (i.e. counts) 
1. the **sampling events** (i.e. information about the observation process)


## Quick Start
As mentioned above, this package is in early development stages and functions are subject to change.
```{r, echo=TRUE, warning=FALSE, message=FALSE, result=FALSE}
# devtools::install_github("trashbirdecology/dubcorms")
library(dubcorms)
```

Current exported functions:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
lsf.str("package:dubcorms")[2:22]
```

