---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval=FALSE,## this is false for development purposes.
  comment = "#>",
  message=FALSE, 
  warning=FALSE, 
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/trashbirdecology/bbsebird/workflows/R-CMD-check/badge.svg)](https://github.com/trashbirdecology/bbsebird/actions)
<!-- badges: end -->

# bbsebird
The purpose of this R package (*likely to undergo a name change...*) is to:

  1. provide a (currently) faster alternative to the R package `auk ` for importing and munging the large eBird datasets*
  1. integrate the BBS and eBird observation datasets for use in JAGS (`rjags`) and `mcgv::jagam()` by binding data to a common, spatial sampling grid
  
> *[\@cboettig](https://github.com/cboettig/) and [\@amstrimas](https://github.com/amstrimas/) are currently developing an `auk` alternative, [`birddb`](https://github.com/cboettig/birddb/). It is likely that, once stable, this R package will use `birddb` as dependency for eBird import and manipulation. For now, however, the functions herein provide a much faster alternative to `auk`.


## Installation
Download development version from GitHub with:
```{r results=FALSE, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
# install.packages("devtools")
remotes::install_github("trashbirdecology/bbsassistant") # to be safe
remotes::install_github("trashbirdecology/bbsebird")
```

## eBird Data Requirements
Prior to using this package, you must have downloaded eBird data. To request and download eBird observations, visit [the eBird website](https://ebird.org/data/download). Credentials are required, and may take up to a few business days for approval, depending on use case. For more information on the eBird data see the eBird website, or visit [the repository for the Cornell Lab of Ornithology's offical R package for munging eBird data,  `auk`](https://github.com/CornellLabofOrnithology/auk/).

When your account is approved, you will gain access to the [eBird Basic Database (EBD)](https://ebird.org/data/download/ebd). This package requires two components of the EBD to be saved to local file:

1. the **observations** (i.e. counts) 
1. the **sampling events** (i.e. information about the observation process)


# DATA MUNGING
## Step 1: Setup
```{r setup}
#explicitly load some packages
pkgs <- c("bbsebird",
          # "bbsAssistant",
          # "reshape2",
          # "stringr",
          # "dplyr",
          # "sf",
          "dplyr"
          )
# install.packages("mapview")
invisible(lapply(pkgs, library, character.only = TRUE))
rm(pkgs)
```

If using this README, this is the only RMD chunk you shoudl have to edit. Most important are where the eBird data and BBS shapefiles are stored (dir.orig.data) and where you wish to save resulting data/models (dir.proj). The latter need not exist -- if it does not exist the package will create the directory for you. 
```{r required-args}
# REQUIRED ARGUMENTS
dir.orig.data  = "C:/Users/jburnett/OneDrive - DOI/bbsebird-testing/" # this will be improved to be more intuitive re: what data? 
dir.proj       = "C:/users/jburnett/OneDrive - DOI/bbsebird-testing/House_Sparrow/"
species             = c("House Sparrow") ## eventually need to add alookup table to ensure species.abbr and speices align.
species.abbr        = c("houspa") # see ebird filename for abbreviation
### this needs improvement as well...e.g. a species lookup table to link common-speci-abbrev across BBS and eBird data...
##bbs arguments
usgs.layer          = "US_BBS_Route-Paths-Snapshot_Taken-Feb-2020" # name of the USGS BBS route shapefile to use
cws.layer           = "ALL_ROUTES" # name of the Canadian (CWS) BBS route shapefile.
##ebird arguments
mmyyyy              = "dec-2021" # the month and year of the eBird data downloads on file

# Strongly suggested but optional args
##general arguments
# dir.proj  = "C:/Users/jburnett/desktop/testing/"


### see bbsAssistant::region_codes
states              = c("us-fl")
countries           = c("US") ## string of  countries Call \code{bbsebird::iso.codes} to find relevant codes for Countries and States/Prov/Territories.
# species             = c("Double-crested Cormorant", "Nannopterum auritum", "phalacrocorax auritum")
# species.abbr        = c("doccor","dcco", "docco")

year.range          = 2008:2019
base.julian.date    = lubridate::ymd(paste0(min(year.range), c("-01-01"))) # used as base date for Julian dates.
crs.target          = 4326 #target CRS for all created spatial layers

##grid arguments
grid.size           = 1.00 # size in decimal degrees (for US/CAN a good est is 1.00dec deg == 111.11km)

##ebird arguments
min.yday            = 91
max.yday            = 245

##JAGS: arguments for customizing the resulting JAGS data list
jagam.args          = list(bs="ds",k=20, family="poisson", sp.prior="log.uniform", diagonalize=TRUE)

## Munge the states and countries indexes for use in dir/proj dir reation
if(!exists("states")) states <- NULL
if(!is.null(states)){regions <- states}else{regions <- countries}
stopifnot(all(tolower(states) %in% tolower(bbsAssistant::region_codes$iso_3166_2)))
```

This chunk will create new environmental variables for project and data directories based on the project directory and data directory specified above.
```{r dirs-set}
# set_proj_shorthand: this will make all directories within a new dir in dir.proj. this is useful for iterating over species/time/space and saving all resulting information in those directories.
subdir.proj <-  set_proj_shorthand(species.abbr, regions, grid.size, year.range)
dirs        <-  dir_spec(dir.orig.data = dir.orig.data,  
                         dir.proj = dir.proj,
                         subdir.proj = subdir.proj) # create and/or specify directories for later use.
```

## Step 2: Make Integrated Data
### Create a spatial sampling grid
The following chunk creates a spatial sampling grid of size grid.size with units defaulting to the units of crs.target.
```{r spatial-grid}
if(is.null(states)){ states.ind <- NULL}else{states.ind<-gsub(x=toupper(states), pattern="-", replacement="")}
study_area <- make_spatial_grid(dir.out = dirs[['dir.spatial.out']],
                          # overwrite=overwrite.grid,
                          states = states.ind,
                          countries = countries,
                          crs.target = crs.target,
                          grid.size = grid.size
                          )
plot(study_area[1])
```

Create the BBS data. This chunk relieson R package \code{bbsAssistant}. The resulting data is aligned with the spatial grid (see above).
```{r bbs-data}
## if the files already exist, don't overwrite unless you've made changes to data specs
if("bbs_obs.rds" %in% list.files(dirs$dir.bbs.out)){bbs_obs <- readRDS(list.files(dirs$dir.bbs.out, "bbs_obs.rds", full.names=TRUE))}else{
bbs_orig <- bbsAssistant::grab_bbs_data(bbs_dir = dirs$dir.bbs.out) 
bbs_obs  <- bbsAssistant::munge_bbs_data(
    bbs_list = bbs_orig,
    states   = states,
    species = species, 
    year.range = year.range)
bbs_obs <- bbsebird:::match_col_names(bbs_obs) # munge column names to mesh with eBird
saveRDS(bbs_obs, paste0(dirs$dir.bbs.out, "/bbs_obs.rds")) # suggest saving data to file for easy access
}
# Overlay BBS and study area / sampling grid
### note, sometimes when running this in a notebook/rmd a random .rdf" path error occurs.
#### I have no clue what this bug is. Just try running it again. See also https://github.com/rstudio/rstudio/issues/6260
if("bbs_spatial.rds" %in% list.files(dirs$dir.bbs.out)){bbs_spatial <- readRDS(list.files(dirs$dir.bbs.out, "bbs_spatial.rds", full.names=TRUE))}else{
bbs_spatial <- make_bbs_spatial(
  df = bbs_obs,
  cws.routes.dir = dirs$cws.routes.dir,
  usgs.routes.dir = dirs$usgs.routes.dir,
  plot.dir = dirs$dir.plots,
  crs.target = crs.target,
  grid = study_area,
  dir.out = dirs$dir.spatial.out, 
  overwrite=TRUE
)
saveRDS(bbs_spatial, paste0(dirs$dir.bbs.out, "/bbs_spatial.rds"))
}
## check out the bbs spatial data to ensure things look ok
plot(bbs_spatial['area']) # cell area
```
Munge the eBird data (must be saved to file):
```{r ebird-data}
## check the specified ebird original data directory for files. 
(fns.ebird    <- id_ebird_files(
  dir.ebird.in = dirs$dir.ebird.in,
  dir.ebird.out = dirs$dir.ebird.out,
  mmyyyy = mmyyyy,
  species = species.abbr,
  states.ind = states
))
stopifnot(length(fns.ebird) > 1)

# Import and munge the desired files
ebird <- munge_ebird_data(
  fns.ebird = fns.ebird,
  species = c(species, species.abbr),
  dir.ebird.out = dirs$dir.ebird.out,
  countries = countries,
  states = states,
  # overwrite = FALSE, ## this function checks for exisiting, munged files iin dir.ebird.out..
  years = year.range
)

# Create spatial ebird
ebird_spatial <- make_ebird_spatial(
  df = ebird,
  crs.target = crs.target,
  grid = study_area,
  overwrite = FALSE, # this fun checks for existing spatial ebird file in dir.spatial.out
  dir.out = dirs$dir.spatial.out 
)
```

## Step 3: Bundle Data for Use in BUGS
Create a list of lists and indexes for use in JAGS or elsewhere. We suggest creating a list using `bundle_data` and subsequently grabbing useful data from there
```{r bundle_data}
bundle <- bundle_data(
  # data
  bbs = bbs_spatial,
  ebird = ebird_spatial,
  grid = study_area,
  # optional args
  # dev.mode = TRUE, ## NED TO FIXBUG:::TRUE creates small output of supplied data sources for rapid model dev/testing
  bf.method = "cubic2D", # method for producing the splines for capturing spatial autocorrelation
  scale.covs = TRUE, 
  fill.cov.nas = 666 # fill missing covariate values (optional)
)
```
 
# BAYESIAN HIERARCHICAL MODELING 
Here, we provide one approach to creating, compiling, and . Additional functionality for compiling models, MCMC samplers, and for sampling from posterior distributions can be found in the R packages nimble and jagsUI or rjags. 

## Step 1: Specify Model
To facilitate modeling using either JAGS or Nimble, we recommend first creating a text file of a standard BUGS model. However, to be compatible within Nimble, the user should not provide any empty indexes (e.g., myMatrix[,] should be myMatrix[1:N, 1:M]). 

We provide some model files in the package:
```{r view.model}

```




