---
title: "0_setup"
params:
  species:  "DOCCOR,DOCCO,DCCO,DCCOR,Double-crested Cormorant,Double Crested Cormorant"
  # dir.orig.data: "C:/Users/aroyle/DOI/Burnett, Jessica L - cormorants/dubcorm-data-backup/"
  dir.orig.data: "C:/Users/jburnett/OneDrive - DOI/research/cormorants/dubcorm-data-backup/"
  dir.proj: "C:/Users/jburnett/documents/github/dubcorms/inst/tutorials/DCCO/"
  countries: "Canada, USA,United States,United States of America" # used to create base maps
  states: "FL, Florida, GA, Georgia" # states/regions of interest
  overwrite.jdat: FALSE
  overwrte.ebird: FALSE # if TRUE, will re-process all eBird data. Do not do this unless you need to re-create the eBird data for some reason. NOTE: If you change a data subsetting parameter (e.g., year.range, grid.size), this will be ignored because the process herein creates a new subdirectory according to parameter specifications. 
  year.range: 2008:2019 # for now, supply a range as integers (will fix later in `clean_params()`)
  mmyyyy: "Oct-2021" # month and year of most recent ebird EBD/samp download in file.
  grid.size: 1.00 #== 1 dec deg
  min.yday: 91  # approximate breeding season day start (day of year)
  max.yday: 245 # approximate breeding season day end   (day of year)
  ebird.protocol: c("Traveling", "Stationary")
  complete.checklists.only: TRUE # ebird
  max.effort.mins: 180 ## arbitrary choice; integer
  max.effort.km:  5 #This is coarse also, typically 5km or less
  max.num.observers: 10 # ebird
  crs.target: 4326 # 4326 == unprojected; 5070=Alberts equal area (used by USGS)
  # include.unid: FALSE ## Whether or not to include UNIDENTIFIED // hybrid species
  get.sunlight: FALSE # if you need to calculate sun/moonrise times set to TRUE (comp. expenesive for eBird data so don't do this unless you realllly want it.) 
---

# Step 0: Setup
```{r chunks, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  # message = TRUE, 
  # echo=FALSE, 
  # warning=FALSE, 
  tidy=TRUE, 
  include=FALSE
)
```

## Libraries
```{r libraries, echo=TRUE}
## Users may run into issues when installing the following packages:
devtools::install_github("ropensci/rnaturalearth", force=FALSE) ## must install from GH -- source has unresolved issues for 2+years (see issue https://github.com/ropensci/rnaturalearthhires/issues/1)
unloadNamespace("dubcorms") # just a precaution to ensure we get most recent dev version of pkg
devtools::install_github("trashbirdecology/dubcorms", force=FALSE)

### specify packages that we call often 
pkgs <- c("dubcorms","bbsAssistant","ggplot2","reshape2",
          "stringr","dplyr","sf", "jagsUI")
invisible(lapply(pkgs, library, character.only = TRUE))
rm(pkgs)
```

## Evaluate Parameters from YAML
Next, use function `dubcorms::eval_params` to clean up the parameters, then assign all  to global environment.  
```{r eval-params, eval=TRUE, include=FALSE}
# eval parameters
params.temp <- eval_params(params) # we cannot save as params--won't allow us to overwrite
list2env(params.temp, envir = environment())
rm(params.temp) # remve temp 
```


## Specify Directories
```{r specs-dirs}
# proj.shorthand: this will make all directories within a new dir in dir.proj. this is useful for iterating over species/time/space and saving all resulting information in those directories.
subdir.proj <- proj.shorthand(species, states, grid.size, year.range)
dirs <- dir_spec(dir.orig.data, dir.proj, subdir.proj) # create series of directories for later use.
list2env(dirs, env=.GlobalEnv)# bind the dir values as named global objects (MUST INCLUDE `env=.GlobalEnv`)
rm(dirs)
```

## Specify and Remove Junk
```{r specify-junk}
# save a list of elements we always want to keep in memory
# we also distinguish between the base argumensts for use later in making metadata tables describing our jags objects. 
args.save <- args.save.base <- c("args.save.base", ls())
```

# Step 1: Check for Existing Files

## Check for existing files
```{r scan-files, include=FALSE}
## If all files do not exist, then something is may have gone awry in a previous session and we should be safe and re-create everything.
list <- scan_files(dir.proj = dir.proj)
list2env(list, .GlobalEnv) # unpack the list objects to environment
rm(list)
```

```{r set-eval-conditions}
# Set eval conditions based on whether the files exist in project directories AND the overwrite.params
j.ind <- ifelse("jdat.rds" %in% list.files(jags.data), 
                ifelse(overwrite.jdat, TRUE, FALSE), # if exists and overwrite is true, then eval=TRUE  
                TRUE # if first cond. is false
                )
e.ind <- ifelse("ebird_spatial.rds" %in% list.files(dir.spatial), 
                ifelse(overwrite.ebird, TRUE, FALSE), # if exists and overwrite is true, then eval=TRUE  
                TRUE # if first cond. is false
                )
b.ind <- ifelse("bbs_spatial.rds" %in% list.files(dir.spatial), 
                ifelse(overwrite.bbs, TRUE, FALSE), # if exists and overwrite is true, then eval=TRUE  
                TRUE # if first cond. is false
                )
g.ind <- ifelse("grid.rds" %in% list.files(dir.spatial), FALSE, TRUE)

```


# Step 2: Munge Data if they DNE
If the files don't exist above, as specified in chunk #set-eval-conditions, the chunks in Step 2 will download, import, and munge where necessary. If you wish to overwrite existing files, delete from file or change evaluation conditions as specified above. 


## Step 2a: Spatial Grid
Load or create the underlying spatial sampling grid, to which the observations will be overlayed. 
```{r spatial-grid, eval=g.ind}
fns <- list.files(dir.spatial.out)
if("grid.rds" %in% fns){grid <- readRDS(paste0(dir.spatial.out, "grid.rds"))}else{
grid <- make_spatial_grid(dir.spatial.out, overwrite=TRUE, states = states)}
# plot(grid[1])
```


## Step 2b: BBS Data
Munging the BBS data shouldn't take too long, regardless spatial extent.
```{r bbs, eval=b.ind}
# If the original data already exists, just import it. Unless you are not using the most recent version (as specified in package `bbsAssistant`), bbs_orig should not change.
fns <- list.files(dir.bbs.out)
if("bbs_orig.rds" %in% fns){bbs_orig <- readRDS(paste0(dir.bbs.out,"bbs_orig.rds"))}else{
  # if DNE, create and import data. 
  bbs_orig <- grab_bbs_data(sb_dir = dir.bbs.out)
}
# Munge the BBS data per specifications in `params`. 
fns <- list.files(dir.bbs.out)
if (!"bbs_obs.rds" %in% tolower(fns)) {
  bbs_obs <-
    munge_bbs(
      list = bbs_orig,
      spp = species,
      states = states,
      keep.stop.level.data = FALSE
    )
  saveRDS(bbs_obs, paste0(dir.bbs.out, "bbs_obs.rds"))
} else{bbs_obs <- readRDS(paste0(dir.bbs.out, "bbs_obs.rds"))}

# Overlay BBS observations and spatial grid
fns.spat <- list.files(dir.spatial.out)
if ("bbs_spatial.rds" %in% tolower(fns.spat)){
  bbs_spatial <- 
    make_bbs_spatial(
    bbs.obs = bbs_obs,
    cws.routes.dir = cws.routes.dir, #location of the CWS BBS routes shapefiles
    usgs.routes.dir = usgs.routes.dir, #location of the USGS BBS routes shapefiles
    crs.target = crs.target,
    routes.keep=unique(bbs_obs$RTENO), 
    grid=grid,
    keep.empty.cells =TRUE,
    plot.dir=dir.plots,
    overwrite=TRUE)
    saveRDS(bbs_spatial, paste0(dir.spatial.out, "bbs_spatial.rds"))
    }else{
  bbs_spatial <- readRDS(paste0(dir.spatial.out, "bbs_spatial.rds"))}
# hist(log(bbs_spatial$RouteTotal))
browseURL(list.files(dir.plots, "bbs", full.names = TRUE))
```

## Step 2c: eBird Data
Munging the eBird will take quite a while, especially if you are running more than a couple of states. This process also requires a significant amount of free RAM. A
```{r ebird, eval=e.ind}
fns.ebird <- id_ebird_files(dir.ebird.in = dir.ebird.in)
fns <- list.files(dir.ebird.out)


# To avoid unnecessary wait times, when overwrite.ebird = FALSE and ebird_spatial.rds already exists, will not run the othre ebird munging processes.
# Import and pre-process the eBird data
ebird_filtered <- }else{ebird_filtered <- filter_ebird_data(
                                    fns.ebird = fns.ebird,
                                    overwrite = FALSE,
                                    dir.ebird.out = dir.ebird.out,
                                    countries = countries,
                                    states = states,
                                    protocol = c("Traveling","Stationary"),
                                    species = species,
                                    max.num.observers=max.num.observers,
                                    years=year.range
                                    )

# Zero-fill the filtered eBird data

ebird_zf <- zerofill_ebird(myList=ebird_filtered, overwrite=FALSE, dir.out=dir.ebird.out)

# Overlay eBird observations and spatial grid


saveRDS(ebird_spatial, file = paste0(dir.spatial.out, "ebird_spatial.rds"))


```



