---
title: "0_setup"
params:
  species:  "DOCCOR,DOCCO,DCCO,DCCOR,Double-crested Cormorant,Double Crested Cormorant"
  # dir.orig.data: "C:/Users/aroyle/DOI/Burnett, Jessica L - cormorants/dubcorm-data-backup/"
  dir.orig.data: "C:/Users/jburnett/OneDrive - DOI/research/cormorants/dubcorm-data-backup/"
  dir.proj: "C:/Users/jburnett/documents/github/dubcorms/inst/tutorials/DCCO/"
  countries: "Canada, USA,United States,United States of America" # used to create base maps
  states: "FL, Florida, GA, Georgia" # states/regions of interest
  overwrite.jdat: FALSE
  overwrite.bbs:  FALSE
  overwrte.ebird: FALSE # if TRUE, will re-process all eBird data. Do not do this unless you need to re-create the eBird data for some reason. NOTE: If you change a data subsetting parameter (e.g., year.range, grid.size), this will be ignored because the process herein creates a new subdirectory according to parameter specifications. 
  year.range: 2008:2019 # for now, supply a range as integers (will fix later in `clean_params()`)
  mmyyyy: "Oct-2021" # month and year of most recent ebird EBD/samp download in file.
  grid.size: 1.00 #== 1 dec deg
  min.yday: 91  # approximate breeding season day start (day of year)
  max.yday: 245 # approximate breeding season day end   (day of year)
  ebird.protocol: c("Traveling", "Stationary")
  complete.checklists.only: TRUE # ebird
  max.effort.mins: 180 ## arbitrary choice; integer
  max.effort.km:  5 #This is coarse also, typically 5km or less
  max.num.observers: 10 # ebird
  crs.target: 4326 # 4326 == unprojected; 5070=Alberts equal area (used by USGS)
  # include.unid: FALSE ## Whether or not to include UNIDENTIFIED // hybrid species
  get.sunlight: FALSE # if you need to calculate sun/moonrise times set to TRUE (comp. expenesive for eBird data so don't do this unless you realllly want it.) 
---

# Step 0: Setup
```{r chunks, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  # message = TRUE, 
  # echo=FALSE, 
  # warning=FALSE, 
  tidy=TRUE, 
  include=FALSE
)
```

## Libraries
```{r libraries, echo=TRUE}
## Users may run into issues when installing the following packages:
devtools::install_github("ropensci/rnaturalearth", force=FALSE) ## must install from GH -- source has unresolved issues for 2+years (see issue https://github.com/ropensci/rnaturalearthhires/issues/1)
unloadNamespace("dubcorms") # just a precaution to ensure we get most recent dev version of pkg
devtools::install_github("trashbirdecology/dubcorms", force=FALSE)

### specify packages that we call often 
pkgs <- c("dubcorms","bbsAssistant","ggplot2","reshape2",
          "stringr","dplyr","sf", "jagsUI")
invisible(lapply(pkgs, library, character.only = TRUE))
rm(pkgs)
```

## Evaluate Parameters from YAML
Next, use function `dubcorms::eval_params` to clean up the parameters, then assign all  to global environment.  
```{r eval-params, eval=TRUE, include=FALSE}
# eval parameters
params.temp <- eval_params(params) # we cannot save as params--won't allow us to overwrite
list2env(params.temp, envir = environment())
rm(params.temp) # remve temp 
```


## Specify Directories
```{r specs-dirs}
# proj.shorthand: this will make all directories within a new dir in dir.proj. this is useful for iterating over species/time/space and saving all resulting information in those directories.
subdir.proj <- proj.shorthand(species, states, grid.size, year.range)
dirs <- dir_spec(dir.orig.data, dir.proj, subdir.proj) # create series of directories for later use.
list2env(dirs, env=.GlobalEnv)# bind the dir values as named global objects (MUST INCLUDE `env=.GlobalEnv`)
rm(dirs)
```

## Specify and Remove Junk
```{r specify-junk}
# DO NOT SKIP
# save a list of elements we always want to keep in memory. 
# later, we use `junk_it()` to build upon this list (and use it to rm() objs not in this list)
args.save <- args.save.base <- c("args.save.base", ls())
```

# Step 1: Check for Existing Files

## Check for existing files
```{r scan-files, include=FALSE}
## If all files do not exist, then something is may have gone awry in a previous session and we should be safe and re-create everything.
list <- scan_files(dir.proj = dir.proj)
list2env(list, .GlobalEnv) # unpack the list objects to environment
rm(list)
```

```{r set-eval-conditions}
j.ind <- ifelse(!exists("jdat"), TRUE, FALSE)
e.ind <- ifelse(!exists("ebird_spatial"), TRUE ,FALSE)
b.ind <- ifelse(!exists("bbs_spatial"), TRUE, FALSE)
g.ind <- ifelse(!exists("grid"), TRUE, FALSE)
```

# Step 2: Data Creation and Basic Munging
If the files don't exist above, as specified in chunk #set-eval-conditions, the chunks in Step 2 will download, import, and munge where necessary. If you wish to overwrite existing files, delete from file or change evaluation conditions as specified above. 

## Step 2a: Spatial Grid
Load or create the underlying spatial sampling grid, to which the observations will be overlayed. 
```{r spatial-grid, eval=g.ind}
if(g.ind) grid <- make_spatial_grid(dir.spatial.out, overwrite=TRUE, states = states)
```

## Step 2b: BBS Data
Munging the BBS data shouldn't take too long, regardless spatial extent.
```{r bbs, eval=b.ind}
# If the original data already exists, just import it. Unless you are not using the most recent version (as specified in package `bbsAssistant`), bbs_orig should not change.
## If b.ind is FALSE then bbs_spatial should already exist. 
# If TRUE, then first ensure bbs_obs,  read in bbs_spatial, otherwise, create or import the bbs observations and create bbs_spatial
fns <- tolower(list.files(dir.bbs.out))
if(b.ind & overwrite.bbs){
  if("bbs_orig.rds" %in% fns){bbs_orig <- readRDS(paste0(dir.bbs.out,"bbs_orig.rds"))}else{
  # if DNE, create and import data. 
  bbs_orig <- grab_bbs_data(sb_dir = dir.bbs.out)}
  
  # If exists on file, load in the observations
  if (!exists("bbs_obs") & "bbs_obs.rds" %in% fns){bbs_obs <- readRDS(paste0(dir.bbs.out, "bbs_obs.rds"))}
  ## go ahead and munge if it still DNE
  if (!exists("bbs_obs")) {
    bbs_obs <-
      munge_bbs(
        list = bbs_orig,
        spp = species,
        states = states,
        keep.stop.level.data = FALSE
      )
    saveRDS(bbs_obs, paste0(dir.bbs.out, "bbs_obs.rds"))}
   }

# Load in or overlay BBS observations and spatial grid
if(!exists("bbs_spatial") | overwrite.bbs){
fns.spat <- list.files(dir.spatial.out)
  bbs_spatial <- 
    make_bbs_spatial(
    bbs.obs = bbs_obs,
    cws.routes.dir = cws.routes.dir, #location of the CWS BBS routes shapefiles
    usgs.routes.dir = usgs.routes.dir, #location of the USGS BBS routes shapefiles
    crs.target = crs.target,
    routes.keep=unique(bbs_obs$RTENO), 
    grid=grid,
    keep.empty.cells =TRUE,
    plot.dir=dir.plots,
    overwrite=TRUE)
    saveRDS(bbs_spatial, paste0(dir.spatial.out, "bbs_spatial.rds"))
    }

browseURL(list.files(dir.plots, "bbs", full.names = TRUE))
```

## Step 2c: eBird Data
Munging the eBird will take quite a while, especially if you are running more than a couple of states. This process also requires a significant amount of free RAM. A
```{r ebird, eval=e.ind}
fns <- list.files(dir.ebird.out)
# To avoid unnecessary wait times, when overwrite.ebird = FALSE and ebird_spatial.rds already exists, will not run the othre ebird munging processes.
# Import and pre-process the eBird data

# If e.ind=FALSE, just import the ebird_spatial data object. If true, will take a while to re-do filtering etc.
if(!e.ind & !exists("ebird_spatial")){ebird_spatial <- readRDS(paste0(dir.spatial.out, "ebird_spatial.rds"))}

if(e.ind){
  # import existing filtered ebird if it exists:
  if("ebird_filtered.rds" %in% fns){ebird_filtered <- readRDS(paste0(dir.ebird.out, "ebird_filtered.rds"))}else{
    gc() # just to maximize space avail
    fns.ebird <- id_ebird_files(dir.ebird.in = dir.ebird.in)
    ebird_filtered <- filter_ebird_data(
        fns.ebird = fns.ebird, 
        overwrite = FALSE,
        dir.ebird.out = dir.ebird.out,
        countries = countries,
        states = states,
        protocol = c("Traveling", "Stationary"),
        species = species,
        max.num.observers = max.num.observers,
        years = year.range
    )
    saveRDS(ebird_filtered, paste0(dir.ebird.out,  "ebird_filtered.rds"))
  } #end ebird_filtered
    
   if("ebird_zf.rds" %in% fns){ebird_zf <- readRDS(paste0(dir.ebird.out, "ebird_zf.rds"))}else{
     # Zero-fill the filtered eBird data
    ebird_zf <- zerofill_ebird(myList=ebird_filtered, overwrite=FALSE, dir.out=dir.ebird.out)
  saveRDS(ebird_zf, paste0(dir.ebird.out,  "ebird_zf.rds"))
   } # end ebird_zf
  
    # Overlay eBird observations and spatial grid 
    ebird_spatial <- make_ebird_spatial(df=ebird_zf, crs.target = crs.target, grid = grid)
    saveRDS(ebird_spatial, file = paste0(dir.spatial.out, "ebird_spatial.rds"))
}

args.save <- junk_it(args.save, "ebird_spatial")

```



